% !TEX root = ../pdf/lsr.tex
% [There are multiple lsr.tex files, but the one in ../pdf is the usual one]


\clearpage
\newpage

\begin{enumerate} 

\item Copyright notice
\begin{enumerate}
 \item  \copyright\ 2017 Daniel Joseph Navarro, All rights reserved.
 \item International Standard Book Number 978-1-326-18972-3
 \item This material is subject to copyright. The copyright of this material, including, but not limited to, the text, photographs, images, software  (`the Material') is owned by Daniel Joseph Navarro (`the Author'). 
 %music
 %audio material, video material and audio-visual material
 %[Note to Dan – review / delete material list as appropriate]
 \item Except as specifically prescribed by the Copyright Act 1968, no part of the Material may in any form or by any means (electronic, mechanical, microcopying, photocopying, recording or otherwise) be reproduced, stored in a retrieval system or transmitted without the Author's prior written permission.
\item To avoid any doubt -- except as noted in paragraph 4(a) -- the Material must not be, without limitation, edited, changed, transformed, published, republished, sold, distributed, redistributed, broadcast, posted on the internet, compiled, shown or played in public (in any form or media) without the Author's prior written permission. %The Material may only be used for the purpose of providing feedback or other commentary to the Author.
\item The Author asserts his Moral Rights (as defined by the Copyright Act 1968) in the Material.
\end{enumerate}

\item Intellectual property rights
\begin{enumerate}
\item `Intellectual Property' for the purposes of paragraph 2(b), means ``all copyright and all rights in relation to inventions, registered and unregistered trademarks (including service marks), registered and unregistered designs, confidential information and circuit layouts, and any other rights resulting from intellectual activity in the industrial, scientific, literary and artistic fields recognised in domestic law and anywhere in the world''
%\item ‘Source Materials’ for the purposes of paragraphs 1(b) and 2(c) means “the source code of, and all relevant documentation, notes and other materials required to support any software supplied in connection with the Materials”.
\item All Intellectual Property rights in the Material are owned by the Author. No licence or any other rights are granted to any other person in respect of the Intellectual Property contained in the Materials in Australia or anywhere else in the world.
\end{enumerate}

\item No warranty
\begin{enumerate}
\item The Author makes no warranty or representation that the Materials are correct, accurate, current, reliable, complete, or fit for any particular purpose at all and the Author expressly disclaims any other warranties, express or implied either in fact or at law, to the extent permitted by law.
\item The user accepts sole responsibility and risk associated with the use of the Material. In no event will the Author be liable for any loss or damage including special, indirect or consequential damage, suffered by any person, resulting from or in connection with the Author's provision of the Material.
\end{enumerate}

\item Preservation of GPL rights for \R\ code
\begin{enumerate}
\item No terms in this notice shall be construed as implying a limitation on the software distribution rights granted by the GPL licences under which \R\ is licensed. 
\item To avoid ambiguity, paragraph 4(a) means means that all \R\ source code reproduced in the Materials but not written by the Author retains the original distribution rights. In addition, it is the intention of the Author that the ``lsr'' \R\ package with which this book is associated be treated as a distinct work from these Materials. The lsr package is freely available, and is distributed under the GPL. The Materials are not.
\end{enumerate}
\end{enumerate}